---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output:
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r echo=FALSE, include=FALSE, message=FALSE}
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)
library(knitr) # for kable
library(caret) # creating test and training sets
library(rpart) # regression and classification trees
library(pROC) # roc plot
library(ggplot2)
library(rattle) #for fancyRpartPlot
library(randomForest)
# for fixing the random generator version and seed
library(recipes)
library(embed)
library(tidymodels)
r.version <- "4.3.2"
seed.val <- 123456
```

# Overview

There are many factors that determine the salary of a job. There are aspects that the job requires and qualities that the applicant can uniquely bring. Knowing how each of these predictors affect the salary of a job is of major interest to employers and those looking to be employed as they help them gauge the market.

The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?


# Data Description and Processing
The dataset used in this analysis was come from Kaggle[1]. We categorized the variables in the dataset as personal and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.


Personal Variables: \
Experience years:Number of years of professional experience. Numeric \
Education level:Highest level of education completed. Categorical \
Skills count: Number of technical or professional skills. Numeric \ 
Certifications Number of professional certifications. Numeric \


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer). Categorical \
Industry: Industry sector where the job belongs. Categorical \
Company size: Size of the company (small, medium, large). Categorical \
Location:Job location or region. Categorical \
Remote Work:Whether the job allows remote work. Categorical \

Outcome Variable: 
Salary: Annual salary of the employee. Numeric \

The dataset was check for missing values. Of the original 250,000 observations, we found 0 observation that contained missing value. 



```{r echo=FALSE}
js.df<- read.csv("job_salary_prediction_dataset.csv")
```


# Data Exploration 

In figure 1, the distribution of the outcome variable, salary, shows a mostly symmetrical shape that is mildly skewed to the right.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Salary distribution."}
hist(js.df$salary/1000000, freq = FALSE, # freq = FALSE for density scale on y-axis
     main = "", # turn off the title as using the fig.cap
     xlab = "Salary (in millions)",
     ylab = "Density",
     ylim = c(0, 11)
     )
lines(density(js.df$salary/1000000), col = "blue", lwd = 2)
```

Since the median is lower than the mean, the data is slightly right-skewed.

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$salary
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Salary descriptive statistics")
```


\newpage
The distribution of the histogram is a uniform distribution with a non-obvious outlier at the low end 
```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Experience Years."}
hist(js.df$experience_years, 
     freq = FALSE, 
     main = "Experience Year",
     yaxt = 'n', # Hide the default axis
     xlab = "Experience Year",
     ylab = "Percentage (%)",
     breaks = 20)

# Create a custom axis that multiplies density by 100
axis(2, at = seq(0, 0.1, by = 0.02), labels = seq(0, 10, by = 2))
lines(density(js.df$experience_years), col = "blue", lwd = 2)
```




Since the distribution of the graph is uniform distribution, the mean and median are nearly the same with a 0.01 difference. 
```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$experience_years
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Experience descriptive statistics")
```


According to the data, the mean salary for entry-level positions is approximately $30,000 less than that of experienced roles. This observation is based solely on years of experience and does not account for variables such as job title
```{r echo=FALSE}
# 1. Calculate and store the means
mean_entry <- mean(js.df$salary[js.df$experience_years == 0], na.rm = TRUE)
mean_exp   <- mean(js.df$salary[js.df$experience_years != 0], na.rm = TRUE)

# 2. Create a data frame for the table
comparison_table <- data.frame(
  Experience_Level = c("Entry Level (0 Years)", "Experienced (1+ Years)"),
  Average_Salary = c(mean_entry, mean_exp)
)

# 3. Generate the pretty table
kable(comparison_table, 
      col.names = c("Experience Group", "Mean Salary"),
      digits = 2,
      format.args = list(big.mark = ","),
      caption = "Salary Comparison: Entry Level vs. Experienced")
```


\newpage
The line plot demonstrates a positive correlation between years of experience and salary. This upward trend indicates that as professional experience increases, compensation levels tend to rise accordingly
```{r echo=FALSE, fig.cap="Average salary by years of experience."}
# Calculate means using aggregate
summary_base <- aggregate(salary ~ experience_years, data = js.df, mean)

# Plot
plot(summary_base$experience_years, summary_base$salary, 
     type = "b", # "b" means both lines and points
     col = "blue",
     xlab = "Experience", ylab = "Salary",
     main = "Salary Trend")
```


\newpage
The distribution of the Job Title is uniform distributed, nearly all of the job title has the similar percentage. 
```{r echo=FALSE}
# 1. Create a frequency table
job_counts <- table(js.df$job_title)

# 2. Convert to percentages
job_percent <- prop.table(job_counts) * 100

# 3. Sort (optional)
job_percent <- sort(job_percent, decreasing = TRUE)

# 4. Create the bar plot
barplot(job_percent, 
        main = "Job Title Distribution", 
        ylab = "Frequency (%)",   # Labels the y-axis
        col = "skyblue", 
        las = 2, 
        cex.names = 0.7)
```


Due to the uniform distribution of the data, the frequency and percentage for each category are nearly equal, with every job representing a consistent share of approximately 8%
```{r echo=FALSE}
# 1. Get raw counts
counts <- table(js.df$job_title)

# 2. Get percentages (rounded to 2 decimal places for neatness)
percentages <- round(prop.table(counts) * 100, 2)

# 3. Combine into a data frame
job_summary <- data.frame(
  Job_Level = names(counts),
  Frequency = as.numeric(counts),
  Percentage = as.numeric(percentages)
)

# 4. Sort by highest frequency
job_summary <- job_summary[order(-job_summary$Frequency), ]

kable(job_summary, col.names = c("Job Title", "Frequency", "Percentage (%)"),
      caption = "Job title frequency and percentage.")
```


\newpage
# Modeling

We fitted a target encoded linear regression model and random forest on all of the factors that contribute to a job's salary. 

## Hypothesis

We hypothesize that job-based factors, specifically job title and location, will be stronger predictors of salary than personal factors such as experience years and education level.


## Model 1 

We fitted a linear regression model using all the predictors, with salary as the predicted variable.The estimates for the coefficients, standard errors and significance results are displayed below:
```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

split      <- initial_split(js.df, prop = 0.75, strata = salary)
train.data <- training(split)
test.data  <- testing(split)

te.recipe <- recipe(salary ~ ., data = train.data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location,  outcome = vars(salary)) %>%
  step_lencode_glm(industry,  outcome = vars(salary))

mod1     <- linear_reg(engine = "lm")
wf.te    <- workflow() %>% add_recipe(te.recipe) %>% add_model(mod1)
mod1.fit <- fit(wf.te, data = train.data)

prepped.rec <- extract_recipe(mod1.fit)
lm.engine   <- extract_fit_engine(mod1.fit)
```

```{r echo=FALSE}
coef.vals <- coef(summary(lm.engine))
kable(coef.vals, digits = 4,
      caption = "Model 1 coefficient estimates, standard errors, and significance.")
```

```{r echo=FALSE, results='asis'}
adjrsq <- summary(lm.engine)$adj.r.squared
cat(sprintf("The adjusted R-squared for this model was %.4f, meaning the model explains approximately %.1f%% of the variance in salary.",
            adjrsq, adjrsq * 100))
```
```{r echo=FALSE}
overall.stats <- glance(extract_fit_engine(mod1.fit))

f.tbl <- data.frame(
  `F Statistic` = round(overall.stats$statistic, 2),
  `Numerator df` = overall.stats$df,
  `Denominator df` = overall.stats$df.residual,
  `p-value` = format(overall.stats$p.value, scientific = TRUE, digits = 3),
  check.names = FALSE
)

kable(f.tbl, caption = "Overall F-test results for Model 1.")
```

The F-test evaluates whether the model as a whole explains a statistically significant portion of the variation in salary. The result in Table 6 confirms that the predictors have a significant linear relationship with salary. Given the large sample size, this result is expected, with this many observations, even very small effects will produce a significant F-statistic. The adjusted R-squared of 0.9633 is therefore the more informative measure of practical model quality, indicating the model explains approximately 96.3% of the variance in salary.

To compare the relative importance of each predictor, we ranked them by the absolute value of their coefficient, which reflects the magnitude of their effect on salary.

```{r echo=FALSE}
# Build importance table — exclude intercept
coef.tbl <- as.data.frame(coef.vals)
coef.tbl$Predictor <- rownames(coef.tbl)
coef.tbl <- coef.tbl[coef.tbl$Predictor != "(Intercept)", ]

# Clean up predictor labels for the report
coef.tbl$Predictor <- gsub("education_levelDiploma",    "Education: Diploma vs. Bachelor",    coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelHigh School","Education: High School vs. Bachelor", coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelMaster",     "Education: Master vs. Bachelor",      coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelPhD",        "Education: PhD vs. Bachelor",         coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeLarge",         "Company Size: Large vs. Enterprise",  coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeMedium",        "Company Size: Medium vs. Enterprise", coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeSmall",         "Company Size: Small vs. Enterprise",  coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeStartup",       "Company Size: Startup vs. Enterprise",coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("remote_workNo",             "Remote Work: No vs. Hybrid",          coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("remote_workYes",            "Remote Work: Yes vs. Hybrid",         coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("experience_years",          "Experience Years",                    coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("skills_count",              "Skills Count",                        coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("certifications",            "Certifications",                      coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^job_title$",               "Job Title (target encoded)",          coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^location$",                "Location (target encoded)",           coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^industry$",                "Industry (target encoded)",           coef.tbl$Predictor)

imp.tbl <- coef.tbl %>%
  arrange(desc(abs(Estimate))) %>%
  mutate(Rank = row_number(), Coefficient = round(Estimate, 2)) %>%
  select(Rank, Predictor, Coefficient)

kable(imp.tbl,
      caption = "Predictors ranked by importance (absolute coefficient value) for Model 1.")
```

The target-encoded job title and location predictors ranked most important, each with a coefficient close to 1. This means that for every dollar increase in a category's average salary, the model predicts approximately one additional dollar in individual salary — confirming that the role and location are the most significant. Industry ranked near the bottom: the ten industry categories in this data encode to nearly identical average salaries, so the predictor contributes almost no salary signal to the model.

Among the directly interpreted predictors, company size showed the largest effects. Employees at startups earned approximately \$42,000 less on average than those at enterprise-sized companies, holding all other factors constant. Small and medium companies showed similar but smaller gaps. Education level showed meaningful differences, with PhD holders earning roughly \$21,000 more than those with a bachelor's degree. Experience years, skills count, and certifications all had positive but more modest effects.

These results support the hypothesis — job title and location ranked as the 
two most important predictors, confirming that job-based factors dominate 
salary prediction in the linear model.

The following tables show the average actual salary and average predicted salary for each level of the three target-encoded predictors. These allow the reader to interpret the encoded values in plain terms.

```{r echo=FALSE}
inspect.train.tbl <- train.data %>%
  transmute(salary, job_title, location, industry,
            experience_years, education_level, skills_count,
            certifications, company_size, remote_work) %>%
  bind_cols(
    bake(prepped.rec, new_data = train.data) %>%
      transmute(job_title_te = job_title,
                location_te  = location,
                industry_te  = industry),
    predict(mod1.fit, new_data = train.data)
  )
```

```{r echo=FALSE}
job.avg.tbl <- inspect.train.tbl %>%
  group_by(job_title) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),    big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),     big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(job_title_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(`Job Title` = job_title)

kable(job.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by job title.")
```

```{r echo=FALSE}
loc.avg.tbl <- inspect.train.tbl %>%
  group_by(location) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(location_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Location = location)

kable(loc.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by location.")
```

```{r echo=FALSE}
ind.avg.tbl <- inspect.train.tbl %>%
  group_by(industry) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(industry_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Industry = industry)

kable(ind.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by industry.")
```
\newpage
### Evaluation of Model 1 Fit

We checked for multicollinearity among the predictors by calculating the variance inflation factor (VIF) for each.

```{r echo=FALSE}
vif.res <- as.data.frame(vif(lm.engine))
vif.res <- tibble::rownames_to_column(vif.res, var = "Predictor")
kable(vif.res, digits = 2,
      caption = "Generalized variance inflation factors (GVIF) for Model 1 predictors.")
```

All VIF values are small, indicating no meaningful multicollinearity among the predictors.

\newpage

```{r echo=FALSE, out.width="80%", out.height="35%", fig.align='center', fig.cap="Residuals vs. fitted values and Q-Q plot for Model 1."}
par(mfrow = c(1, 2))
plot(lm.engine, which = 1)
plot(lm.engine, which = 2)
par(mfrow = c(1, 1))
```

The residuals vs. fitted plot shows a U-shaped curve rather than a random scatter around zero. The model tends to underpredict salaries at the low and high ends of the fitted range and slightly overpredict in the middle. This suggests the relationship between some predictors and salary is not purely linear. In particular, the effect of experience years on salary likely follows a curve, with larger gains at earlier career stages that level off over time. A linear model cannot capture this shape, so the pattern appears in the residuals. This is a known limitation of linear regression when predictor effects are non-linear, and it motivates the use of a more flexible model such as the random forest in Model 2.

```{r echo=FALSE}
ct <- coeftest(lm.engine, vcov. = vcovHC(lm.engine, type = "HC3"))
kable(ct[, 1:4], digits = 4,
      caption = "Heteroskedasticity-robust (HC3) standard errors for Model 1.")
```

```{r echo=FALSE}
bpt <- bptest(lm.engine)
bps  <- round(as.numeric(bpt$statistic), 4)
degf <- as.numeric(bpt$parameter)
pv <- ifelse(bpt$p.value < 0.0001, "< 0.0001", round(bpt$p.value, 4))
bp.tbl <- tibble(bps, degf, pv)
kable(bp.tbl, col.names = c("BP Statistic", "df", "p-value"),
      caption = "Breusch-Pagan test results for Model 1.")
```

The Breusch-Pagan test returned a highly significant result, indicating the 
presence of heteroskedasticity. However, after applying HC3 robust standard 
errors, the significance levels of all predictors were unchanged, confirming that the coefficient estimates remain reliable despite the heteroskedasticity.

## Model 2 

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

small.sample <- train.data[sample(nrow(train.data), nrow(train.data) * 0.1), ]
salary.rf <- randomForest(salary ~ ., data = small.sample, mtry = 3, ntree = 500, importance = TRUE)
```

```{r echo=FALSE}
imp.vals <- importance(salary.rf, type = 1)  # type=1: %IncMSE
imp.df <- data.frame(
  Predictor  = rownames(imp.vals),
  Importance = round(imp.vals[, 1], 2)
)
imp.df <- imp.df[order(-imp.df$Importance), ]
imp.df$Rank <- seq_len(nrow(imp.df))
imp.df <- imp.df[, c("Rank", "Predictor", "Importance")]

kable(imp.df, row.names = FALSE,
      col.names = c("Rank", "Predictor", "% Increase in MSE"),
      caption = "Predictor importance for Model 2 (random forest), ranked by percent increase in mean squared error.")
```

These results partially contradict the hypothesis. While location and job 
title remain meaningful predictors, experience years, company size, and 
education level all ranked above them, suggesting personal factors carry 
more salary signal than predicted when non-linear effects are accounted for.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="Predictor importance plot for Model 2 (random forest)."}
varImpPlot(salary.rf, type = 1, main = "")
```

Experience years ranked as the most important predictor, followed by company 
size and education level. Location came fourth and job title fifth. This 
differs from Model 1, where job title and location ranked highest through 
their target-encoded coefficients. The random forest ranks personal factors 
more highly because it can capture non-linear salary increases across 
experience levels and differences between education tiers that a linear model can only partially represent.

Industry ranked last with a negative importance value of -0.08, meaning 
removing it from the model would not meaningfully change prediction accuracy. This is consistent with Model 1's finding that industry contributes almost no salary signal in this data.

### Model 2 Fit

```{r echo=FALSE, results='asis'}
rf.preds <- predict(salary.rf, newdata = test.data)
rf.rmse  <- sqrt(mean((rf.preds - test.data$salary)^2))
rf.mae   <- mean(abs(rf.preds - test.data$salary))
cat(sprintf("Evaluated on the test set, the random forest produced a root mean squared error (RMSE) of %s and a mean absolute error (MAE) of %s.",
            format(round(rf.rmse), big.mark = ","),
            format(round(rf.mae),  big.mark = ",")))
```


```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="Predicted vs. actual salary for Model 2 (random forest) on the test set."}
plot(test.data$salary, rf.preds,
     xlab = "Actual Salary",
     ylab = "Predicted Salary",
     pch  = 20, col = rgb(0, 0, 0, 0.05))
abline(0, 1, col = "blue", lwd = 2)
```

The predicted vs. actual salary plot in figure 5 shows the predictions tracking closely along the diagonal, indicating the model is making accurate salary estimates across the full range of the outcome. There is slightly more spread at the higher salary values, suggesting the model has more difficulty predicting the largest salaries precisely.

## Comparison of Models 1 and 2

```{r echo=FALSE}
# Collect test-set metrics for both models
m1.preds <- predict(mod1.fit, new_data = test.data)$.pred
m1.rmse  <- sqrt(mean((m1.preds - test.data$salary)^2))
m1.mae   <- mean(abs(m1.preds - test.data$salary))

comp.tbl <- data.frame(
  Model = c("Model 1: Linear Regression", "Model 2: Random Forest"),
  RMSE  = c(round(m1.rmse), round(rf.rmse)),
  MAE   = c(round(m1.mae),  round(rf.mae))
)
kable(comp.tbl, col.names = c("Model", "RMSE", "MAE"),
      caption = "Test-set prediction error comparison for Models 1 and 2.")
```

Model 1 has a lower RMSE than Model 2, which is somewhat surprising given that random forests are generally more flexible. The likely explanation is each predictor contributes independently to salary, which is exactly the structure linear regression is designed for. The target encoding in Model 1 also gave the linear model an efficient way to capture the salary signal in high-cardinality categoricals, leveling the playing field. The random forest, trained on only 10% of the available data due to computational constraints, did not have access to the full signal that the linear model used.

Both models agree that industry contributes almost no salary signal. Beyond 
that, the models tell a somewhat different story. The linear regression ranked job title and location highest, while the random forest ranked experience years and company size above them. This suggests that the salary effect of experience and education is non-linear — patterns the linear model can only partially capture, but which the random forest detects more clearly.

## Model Selection

In the initial model, two predictors showed negligible contributions to salary prediction. Industry ranked last in importance with a coefficient near zero, and its encoded values fell within a narrow range across all ten categories. Remote work showed a similarly weak effect — the difference in average salary between hybrid and non-remote positions was under $40, and the fully remote premium was modest relative to other predictors. Both predictors were removed and the model was refit on the remaining eight predictors.

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

te.recipe2 <- recipe(salary ~ ., data = train.data) %>%
  step_rm(industry, remote_work) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location,  outcome = vars(salary))

wf.te2   <- workflow() %>% add_recipe(te.recipe2) %>% add_model(mod1)
mod2.fit <- fit(wf.te2, data = train.data)

prepped.rec2 <- extract_recipe(mod2.fit)
lm.engine2   <- extract_fit_engine(mod2.fit)
```

```{r echo=FALSE}
coef.vals2 <- coef(summary(lm.engine2))
kable(coef.vals2, digits = 4,
      caption = "Revised Model 1 coefficient estimates, standard errors, and significance.")
```

```{r echo=FALSE}
coef.tbl2 <- as.data.frame(coef.vals2)
coef.tbl2$Predictor <- rownames(coef.tbl2)
coef.tbl2 <- coef.tbl2[coef.tbl2$Predictor != "(Intercept)", ]

coef.tbl2$Predictor <- gsub("education_levelDiploma",    "Education: Diploma vs. Bachelor",     coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelHigh School","Education: High School vs. Bachelor",  coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelMaster",     "Education: Master vs. Bachelor",       coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelPhD",        "Education: PhD vs. Bachelor",          coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeLarge",         "Company Size: Large vs. Enterprise",   coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeMedium",        "Company Size: Medium vs. Enterprise",  coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeSmall",         "Company Size: Small vs. Enterprise",   coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeStartup",       "Company Size: Startup vs. Enterprise", coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("experience_years",          "Experience Years",                     coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("skills_count",              "Skills Count",                         coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("certifications",            "Certifications",                       coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("^job_title$",               "Job Title (target encoded)",           coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("^location$",                "Location (target encoded)",            coef.tbl2$Predictor)

imp.tbl2 <- coef.tbl2 %>%
  arrange(desc(abs(Estimate))) %>%
  mutate(Rank = row_number(), Coefficient = round(Estimate, 2)) %>%
  select(Rank, Predictor, Coefficient)

kable(imp.tbl2,
      caption = "Predictors ranked by importance for the revised Model 1.")
```

The revised model produced an adjusted R-squared of 0.9635, compared to
0.9633 in the original model, which is a negligible difference confirming that neither removed predictor was contributing meaningful predictive power.
The test-set RMSE of 7,157 was effectively identical to the original model.

The importance rankings were stable after removal: job title and location
were still at the top, continuing to support the hypothesis that job-based
factors are the most significant to salary. The coefficients for company size, education level, experience years, skills count, and certifications were all the same, indicating that industry and remote work were not masking or distorting any of the remaining predictors' effects.

# References
[1] https://www.kaggle.com/datasets/nalisha/job-salary-prediction-dataset
