---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output:
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r echo=FALSE, include=FALSE, message=FALSE}
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)
library(knitr) # for kable
library(caret) # creating test and training sets
library(rpart) # regression and classification trees
library(pROC) # roc plot
library(ggplot2)
library(rattle) #for fancyRpartPlot
library(randomForest)
# for fixing the random generator version and seed
library(recipes)
library(embed)
library(tidymodels)
r.version <- "4.3.2"
seed.val <- 123456
```

# Overview

There are many factors that determine the salary of a job. There are aspects that the job requires and qualities that the applicant can uniquely bring. Knowing how each of these predictors affect the salary of a job is of major interest to employers and those looking to be employed as they help them gauge the market.

The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?


# Data Description and Processing
The dataset used in this analysis was come from Kaggle[1]. We categorized the variables in the dataset as personal and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.


Personal Variables: \
Experience years:Number of years of professional experience. Numeric \
Education level:Highest level of education completed. Categorical \
Skills count: Number of technical or professional skills. Numeric \
Certifications: Number of professional certifications. Numeric \


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer). Categorical \
Industry: Industry sector where the job belongs. Categorical \
Company size: Size of the company (small, medium, large). Categorical \
Location: Job location or region. Categorical \
Remote Work:Whether the job allows remote work. Categorical \

Outcome Variable: 
Salary: Annual salary of the employee. Numeric \

The dataset was check for missing values. Of the original 250,000 observations, we found 0 observation that contained missing value. 



```{r echo=FALSE}
js.df<- read.csv("job_salary_prediction_dataset.csv")
```


# Data Exploration 

In figure 1, the distribution of the outcome variable, salary, shows a mostly symmetrical shape that is mildly skewed to the right.

```{r echo=FALSE, out.width="65%", out.height="28%", fig.align = 'center', fig.cap="Salary Density"}
hist(js.df$salary/1000000, freq = FALSE, # freq = FALSE for density scale on y-axis
     main = "",
     xlab = "Salary (in millions)",
     ylab = "Density",
     ylim = c(0, 11)
     )
lines(density(js.df$salary/1000000), col = "blue", lwd = 2)
```

Since the median is lower than the mean, the data is slightly right-skewed.

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$salary
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Salary descriptive statistics")
```


The distribution of the histogram is a uniform distribution with a non-obvious outlier at the low end 
```{r echo=FALSE, out.width="65%", out.height="28%", fig.align = 'center', fig.cap="Years of Experience and Percentage"}
exp_bins <- cut(js.df$experience_years, breaks = seq(0, 20, by = 2), include.lowest = TRUE)
exp_counts <- table(exp_bins)
exp_percent <- round(prop.table(exp_counts) * 100, 2)

par(mar = c(6, 4, 2, 1))
barplot(exp_percent,
        main = "",
        xlab = "Years of Experience",
        ylab = "Percentage (%)",
        col = "skyblue",
        las = 1,
        names.arg = c("0-2","2-4","4-6","6-8","8-10","10-12","12-14","14-16","16-18","18-20"),
        cex.names = 0.8,
        cex.axis = 0.8)
```

\newpage

The table below illustrates the frequency and percentage of each year of experience.
```{r echo=FALSE}
exp_bins <- cut(js.df$experience_years, breaks = seq(0, 20, by = 2), include.lowest = TRUE)
exp_counts <- table(exp_bins)
exp_percent <- round(prop.table(exp_counts) * 100, 2)

exp_summary <- data.frame(
  Range = names(exp_counts),
  Frequency = as.numeric(exp_counts),
  Percentage = as.numeric(exp_percent)
)

kable(exp_summary,
      col.names = c("Range of Experience", "Frequency", "Percentage (%)"),
      caption = "Years of Experience, Frequency, and Percentage.")
```


According to the data, the mean salary for entry-level positions is approximately $30,000 less than that of experienced roles. This observation is based solely on years of experience and does not account for variables such as job title
```{r echo=FALSE}
# 1. Calculate and store the means
mean_entry <- mean(js.df$salary[js.df$experience_years == 0], na.rm = TRUE)
mean_exp   <- mean(js.df$salary[js.df$experience_years != 0], na.rm = TRUE)

# 2. Create a data frame for the table
comparison_table <- data.frame(
  Experience_Level = c("Entry Level (0 Years)", "Experienced (1+ Years)"),
  Average_Salary = c(mean_entry, mean_exp)
)

# 3. Generate the pretty table
kable(comparison_table, 
      col.names = c("Experience Group", "Mean Salary"),
      digits = 2,
      format.args = list(big.mark = ","),
      caption = "Salary Comparison: Entry Level vs. Experienced")
```


The line plot demonstrates a positive correlation between years of experience and salary. This upward trend indicates that as professional experience increases, compensation levels tend to rise accordingly

```{r echo=FALSE, out.width="65%", out.height="28%", fig.align='center', fig.cap="Average salary by years of experience."}
# Calculate means using aggregate
summary_base <- aggregate(salary ~ experience_years, data = js.df, mean)

# Plot
plot(summary_base$experience_years, summary_base$salary, 
     type = "b", # "b" means both lines and points
     main = "",
     col = "blue",
     xlab = "Experience", ylab = "Salary")
```


\newpage
The distribution of the Job Title is uniform distributed, nearly all of the job title has the similar percentage. 
```{r echo=FALSE, fig.align="center", fig.cap="Job Title Distribution"}
job_counts <- table(js.df$job_title)

job_percent <- prop.table(job_counts) * 100

job_percent <- sort(job_percent, decreasing = TRUE)

par(mar = c(10, 4, 2, 1))

barplot(job_percent, 
        main = "", 
        ylab = "Frequency (%)",   
        col = "skyblue", 
        las = 2, 
        cex.names = 0.7, cex.axis =0.8)
```


Due to the uniform distribution of the data, the frequency and percentage for each category are nearly equal, with every job representing a consistent share of approximately 8%
```{r echo=FALSE}
# 1. Get raw counts
counts <- table(js.df$job_title)

# 2. Get percentages (rounded to 2 decimal places for neatness)
percentages <- round(prop.table(counts) * 100, 2)

# 3. Combine into a data frame
job_summary <- data.frame(
  Job_Level = names(counts),
  Frequency = as.numeric(counts),
  Percentage = as.numeric(percentages)
)

# 4. Sort by highest frequency
job_summary <- job_summary[order(-job_summary$Frequency), ]

kable(job_summary, col.names = c("Job Title", "Frequency", "Percentage (%)"),
      caption = "Job title frequency and percentage.")
```


\newpage
# Modeling

We fitted a target encoded linear regression model and random forest on all of the factors that contribute to a job's salary. 

## Hypothesis

We hypothesize that job-based factors, specifically job title and location, will be stronger predictors of salary than personal factors such as experience years and education level.


## Model 1 

We fitted a linear regression model using all the predictors, with salary as the predicted variable.The estimates for the coefficients, standard errors and significance results are displayed below:
```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

split      <- initial_split(js.df, prop = 0.75, strata = salary)
train.data <- training(split)
test.data  <- testing(split)

te.recipe <- recipe(salary ~ ., data = train.data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location,  outcome = vars(salary)) %>%
  step_lencode_glm(industry,  outcome = vars(salary))

mod1     <- linear_reg(engine = "lm")
wf.te    <- workflow() %>% add_recipe(te.recipe) %>% add_model(mod1)
mod1.fit <- fit(wf.te, data = train.data)

prepped.rec <- extract_recipe(mod1.fit)
lm.engine   <- extract_fit_engine(mod1.fit)
```

```{r echo=FALSE}
coef.vals <- coef(summary(lm.engine))
kable(coef.vals, digits = 4,
      caption = "Model 1 coefficient estimates, standard errors, and significance.")
```

```{r echo=FALSE, results='asis'}
adjrsq <- summary(lm.engine)$adj.r.squared
cat(sprintf("The adjusted R-squared for this model was %.4f, meaning the model explains approximately %.1f%% of the variance in salary.",
            adjrsq, adjrsq * 100))
```



```{r echo=FALSE}
overall.stats <- glance(extract_fit_engine(mod1.fit))

f.tbl <- data.frame(
  `F Statistic` = round(overall.stats$statistic, 2),
  `Numerator df` = overall.stats$df,
  `Denominator df` = overall.stats$df.residual,
  `p-value` = format(overall.stats$p.value, scientific = TRUE, digits = 3),
  check.names = FALSE
)

kable(f.tbl, caption = "Overall F-test results for Model 1.")
```

The F-test evaluates whether the model as a whole explains a statistically significant portion of the variation in salary. The result in Table 6 confirms that the predictors have a significant linear relationship with salary. Given the large sample size, this result is expected, with this many observations, even very small effects will produce a significant F-statistic. The adjusted R-squared of 0.9633 is therefore the more informative measure of practical model quality, indicating the model explains approximately 96.3% of the variance in salary.

\newpage 

To compare the relative importance of each predictor, we ranked them by the absolute value of their coefficient, which reflects the magnitude of their effect on salary.

```{r echo=FALSE}
# Build importance table — exclude intercept
coef.tbl <- as.data.frame(coef.vals)
coef.tbl$Predictor <- rownames(coef.tbl)
coef.tbl <- coef.tbl[coef.tbl$Predictor != "(Intercept)", ]

# Clean up predictor labels for the report
coef.tbl$Predictor <- gsub("education_levelDiploma",    "Education: Diploma vs. Bachelor",    coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelHigh School","Education: High School vs. Bachelor", coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelMaster",     "Education: Master vs. Bachelor",      coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("education_levelPhD",        "Education: PhD vs. Bachelor",         coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeLarge",         "Company Size: Large vs. Enterprise",  coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeMedium",        "Company Size: Medium vs. Enterprise", coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeSmall",         "Company Size: Small vs. Enterprise",  coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("company_sizeStartup",       "Company Size: Startup vs. Enterprise",coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("remote_workNo",             "Remote Work: No vs. Hybrid",          coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("remote_workYes",            "Remote Work: Yes vs. Hybrid",         coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("experience_years",          "Experience Years",                    coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("skills_count",              "Skills Count",                        coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("certifications",            "Certifications",                      coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^job_title$",               "Job Title (target encoded)",          coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^location$",                "Location (target encoded)",           coef.tbl$Predictor)
coef.tbl$Predictor <- gsub("^industry$",                "Industry (target encoded)",           coef.tbl$Predictor)

imp.tbl <- coef.tbl %>%
  arrange(desc(abs(Estimate))) %>%
  mutate(Rank = row_number(), Coefficient = round(Estimate, 2)) %>%
  select(Rank, Predictor, Coefficient)

kable(imp.tbl,
      caption = "Predictors ranked by importance (absolute coefficient value) for Model 1.")
```

Company size showed the largest coefficients in the model, with startup 
employees earning approximately 42,000 less than those at enterprise companies. Job title and location, though ranked lowest by coefficient magnitude, are not directly comparable to the dummy-coded predictors — their coefficients near 1.0 reflect the target encoding scale rather than raw salary impact.The hypothesis is better evaluated through the encoded value tables, which show a 53,000 salary range across job titles and an $84,000 range across locations — larger spreads than any other predictor.

The following tables show the average actual salary and average predicted salary for each level of the three target-encoded predictors. These allow the reader to interpret the encoded values in plain terms.

```{r echo=FALSE}
inspect.train.tbl <- train.data %>%
  transmute(salary, job_title, location, industry,
            experience_years, education_level, skills_count,
            certifications, company_size, remote_work) %>%
  bind_cols(
    bake(prepped.rec, new_data = train.data) %>%
      transmute(job_title_te = job_title,
                location_te  = location,
                industry_te  = industry),
    predict(mod1.fit, new_data = train.data)
  )
```

```{r echo=FALSE}
job.avg.tbl <- inspect.train.tbl %>%
  group_by(job_title) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),    big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),     big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(job_title_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(`Job Title` = job_title)

kable(job.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by job title.")
```

```{r echo=FALSE}
loc.avg.tbl <- inspect.train.tbl %>%
  group_by(location) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(location_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Location = location)

kable(loc.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by location.")
```

```{r echo=FALSE}
ind.avg.tbl <- inspect.train.tbl %>%
  group_by(industry) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(industry_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Industry = industry)

kable(ind.avg.tbl,
      caption = "Average actual salary, predicted salary, and encoded value by industry.")
```



### Evaluation of Model 1 Fit

We checked for multicollinearity among the predictors by calculating the variance inflation factor (VIF) for each.

```{r echo=FALSE}
vif.res <- as.data.frame(vif(lm.engine))
vif.res <- tibble::rownames_to_column(vif.res, var = "Predictor")
kable(vif.res, digits = 2,
      caption = "Generalized variance inflation factors (GVIF) for Model 1 predictors.")
```

All VIF values are small, indicating no meaningful multicollinearity among the predictors.


```{r echo=FALSE, out.width="80%", out.height="35%", fig.align='center', fig.cap="Residuals vs. fitted values and Q-Q plot for Model 1."}
par(mfrow = c(1, 2))
plot(lm.engine, which = 1)
plot(lm.engine, which = 2)
par(mfrow = c(1, 1))
```

The residuals vs. fitted plot shows a U-shaped curve rather than a random scatter around zero. The model tends to underpredict salaries at the low and high ends of the fitted range and slightly overpredict in the middle. This suggests the relationship between some predictors and salary is not purely linear. In particular, the effect of experience years on salary likely follows a curve, with larger gains at earlier career stages that level off over time. A linear model cannot capture this shape, so the pattern appears in the residuals. This is a known limitation of linear regression when predictor effects are non-linear, and it motivates the use of a more flexible model such as the random forest in Model 2.

```{r echo=FALSE}
ct <- coeftest(lm.engine, vcov. = vcovHC(lm.engine, type = "HC3"))
kable(ct[, 1:4], digits = 4,
      caption = "Heteroskedasticity-robust (HC3) standard errors for Model 1.")
```

```{r echo=FALSE}
bpt <- bptest(lm.engine)
bps  <- round(as.numeric(bpt$statistic), 4)
degf <- as.numeric(bpt$parameter)
pv <- ifelse(bpt$p.value < 0.0001, "< 0.0001", round(bpt$p.value, 4))
bp.tbl <- tibble(bps, degf, pv)
kable(bp.tbl, col.names = c("BP Statistic", "df", "p-value"),
      caption = "Breusch-Pagan test results for Model 1.")
```

The Breusch-Pagan test returned a highly significant result, indicating the 
presence of heteroskedasticity. However, after applying HC3 robust standard 
errors, the significance levels of all predictors were unchanged, confirming that the coefficient estimates remain reliable despite the heteroskedasticity.

## Model 2 

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

small.sample <- train.data[sample(nrow(train.data), nrow(train.data) * 0.1), ]
salary.rf <- randomForest(salary ~ ., data = small.sample, mtry = 3, ntree = 500, importance = TRUE)
```

```{r echo=FALSE}
imp.vals <- importance(salary.rf, type = 1)  # type=1: %IncMSE
imp.df <- data.frame(
  Predictor  = rownames(imp.vals),
  Importance = round(imp.vals[, 1], 2)
)
imp.df <- imp.df[order(-imp.df$Importance), ]
imp.df$Rank <- seq_len(nrow(imp.df))
imp.df <- imp.df[, c("Rank", "Predictor", "Importance")]

kable(imp.df, row.names = FALSE,
      col.names = c("Rank", "Predictor", "% Increase in MSE"),
      caption = "Predictor importance for Model 2 (random forest), ranked by percent increase in mean squared error.")
```

These results partially contradict the hypothesis. While location and job 
title remain meaningful predictors, experience years, company size, and 
education level all ranked above them, suggesting personal factors carry 
more salary signal than predicted when non-linear effects are accounted for.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="Predictor importance plot for Model 2 (random forest)."}
varImpPlot(salary.rf, type = 1, main = "")
```

This differs from Model 1, where the encoded value tables showed a $53,000 
salary range across job titles and an $84,000 range across locations, 
suggesting strong job-based salary signals despite their low coefficient ranks.

Industry ranked last with a negative importance value of -0.08, meaning 
removing it from the model would not meaningfully change prediction accuracy. This is consistent with Model 1's finding that industry contributes almost no salary signal in this data.

### Model 2 Fit

```{r echo=FALSE, results='asis'}
rf.preds <- predict(salary.rf, newdata = test.data)
rf.rmse  <- sqrt(mean((rf.preds - test.data$salary)^2))
rf.mae   <- mean(abs(rf.preds - test.data$salary))
cat(sprintf("Evaluated on the test set, the random forest produced a root mean squared error (RMSE) of %s and a mean absolute error (MAE) of %s.",
            format(round(rf.rmse), big.mark = ","),
            format(round(rf.mae),  big.mark = ",")))
```


```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="Predicted vs. actual salary for Model 2 (random forest) on the test set."}
plot(test.data$salary, rf.preds,
     xlab = "Actual Salary",
     ylab = "Predicted Salary",
     pch  = 20, col = rgb(0, 0, 0, 0.05))
abline(0, 1, col = "blue", lwd = 2)
```

The predicted vs. actual salary plot in figure 6 shows the predictions tracking closely along the diagonal, indicating the model is making accurate salary estimates across the full range of the outcome. There is slightly more spread at the higher salary values, suggesting the model has more difficulty predicting the largest salaries precisely.

## Comparison of Models 1 and 2

```{r echo=FALSE}
# Collect test-set metrics for both models
m1.preds <- predict(mod1.fit, new_data = test.data)$.pred
m1.rmse  <- sqrt(mean((m1.preds - test.data$salary)^2))
m1.mae   <- mean(abs(m1.preds - test.data$salary))

comp.tbl <- data.frame(
  Model = c("Model 1: Linear Regression", "Model 2: Random Forest"),
  RMSE  = c(round(m1.rmse), round(rf.rmse)),
  MAE   = c(round(m1.mae),  round(rf.mae))
)
kable(comp.tbl, col.names = c("Model", "RMSE", "MAE"),
      caption = "Test-set prediction error comparison for Models 1 and 2.")
```

Model 1 has a lower RMSE than Model 2, which is somewhat surprising given that random forests are generally more flexible. The likely explanation is each predictor contributes independently to salary, which is exactly the structure linear regression is designed for. The target encoding in Model 1 also gave the linear model an efficient way to capture the salary signal in high-cardinality categoricals, leveling the playing field. It should be noted that this comparison is not fully fair. The random forest, trained on only 10% of the available data due to computational constraints, did not have access to the full signal that the linear model used.


Taking both models together, the hypothesis is partially supported: job title and location carry broad salary ranges, but experience years, company size, and education level are stronger predictors overall when non-linear effects are considered.

## Model Selection

In the initial model, two predictors showed negligible contributions to salary prediction. Industry ranked last in importance with a coefficient near zero, and its encoded values fell within a narrow range across all ten categories. Remote work showed a similarly weak effect — the difference in average salary between hybrid and non-remote positions was under $40, and the fully remote premium was modest relative to other predictors. Both predictors were removed and the model was refit on the remaining eight predictors.

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

te.recipe2 <- recipe(salary ~ ., data = train.data) %>%
  step_rm(industry, remote_work) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location,  outcome = vars(salary))

wf.te2   <- workflow() %>% add_recipe(te.recipe2) %>% add_model(mod1)
mod2.fit <- fit(wf.te2, data = train.data)

prepped.rec2 <- extract_recipe(mod2.fit)
lm.engine2   <- extract_fit_engine(mod2.fit)
```

```{r echo=FALSE}
coef.vals2 <- coef(summary(lm.engine2))
kable(coef.vals2, digits = 4,
      caption = "Revised Model 1 coefficient estimates, standard errors, and significance.")
```

\newpage

```{r echo=FALSE}
coef.tbl2 <- as.data.frame(coef.vals2)
coef.tbl2$Predictor <- rownames(coef.tbl2)
coef.tbl2 <- coef.tbl2[coef.tbl2$Predictor != "(Intercept)", ]

coef.tbl2$Predictor <- gsub("education_levelDiploma",    "Education: Diploma vs. Bachelor",     coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelHigh School","Education: High School vs. Bachelor",  coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelMaster",     "Education: Master vs. Bachelor",       coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("education_levelPhD",        "Education: PhD vs. Bachelor",          coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeLarge",         "Company Size: Large vs. Enterprise",   coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeMedium",        "Company Size: Medium vs. Enterprise",  coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeSmall",         "Company Size: Small vs. Enterprise",   coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("company_sizeStartup",       "Company Size: Startup vs. Enterprise", coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("experience_years",          "Years of Experience",                     coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("skills_count",              "Skills Count",                         coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("certifications",            "Certifications",                       coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("^job_title$",               "Job Title (target encoded)",           coef.tbl2$Predictor)
coef.tbl2$Predictor <- gsub("^location$",                "Location (target encoded)",            coef.tbl2$Predictor)

imp.tbl2 <- coef.tbl2 %>%
  arrange(desc(abs(Estimate))) %>%
  mutate(Rank = row_number(), Coefficient = round(Estimate, 2)) %>%
  select(Rank, Predictor, Coefficient)

kable(imp.tbl2,
      caption = "Predictors ranked by importance for the revised Model 1.")
```

The revised model produced an adjusted R-squared of 0.9635, compared to
0.9633 in the original model, which is a negligible difference confirming that neither removed predictor was contributing meaningful predictive power.
The test-set RMSE of 7,157 was effectively identical to the original model.

The importance rankings were stable after removal. The same predictors that
dominated in the original model continued to show the largest effects, and 
the encoded value ranges for job title and location remained unchanged, 
continuing to support the hypothesis that job-based factors carry significant salary signal.

# Model Prediction

We are using repeated cross validation to do the model prediction. In this model, there are 10 folds repeated 3 times with 80% of the data used for training on each fold. The first model is linear regression, while another model is random forest. Due to the large amount of data, we are randomly select 10% of the sample for our analysis. 
```{r echo = FALSE}
small.sample <- train.data[sample(nrow(train.data), nrow(train.data) * 0.1), ]

ctrl<- trainControl(method = "repeatedcv",
                    number=10,
                    repeats=3,p=0.8)

```



The two tables below shows the training results regarding Linear Regression and Random Forest model

## Model 1 Prediction 
```{r echo =FALSE,comment=""}
model1<- train(salary~ ., data= small.sample, method="lm",trControl= ctrl)

model1_results <- model1$results

kable(model1_results, digits = 2, caption = "Model 1 Training Results")
```




To evaluate how well each model generalizes to unseen data, we used repeated cross-validation with 10 folds repeated 3 times. In each fold the data is splt so that 80% of the data was used ofr training on each fold and the remaining 20% held out for testing This process produces 30 total evaluations, and the RMSE and standard deviation are averaged out across all folds to give a stable estimate of out-of-sample error.

## Model 1 Prediction

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

cv.control <- trainControl(
  method  = "repeatedcv",
  number  = 10,
  repeats = 3,
  p  = 0.80
)

baked.train <- bake(prepped.rec2, new_data = train.data)

lm.cv <- train(
  salary ~ .,
  data      = baked.train,
  method    = "lm",
  trControl = cv.control
)

lm.cv.tbl <- data.frame(
  Model    = "Linear Regression",
  Mean_RMSE = round(lm.cv$results$RMSE, 2),
  SD_RMSE   = round(lm.cv$results$RMSESD, 2),
  Mean_MAE  = round(lm.cv$results$MAE, 2),
  Mean_Rsq  = round(lm.cv$results$Rsquared, 4)
)

kable(lm.cv.tbl,
      col.names = c("Model", "Mean RMSE", "SD RMSE", "Mean MAE", "Mean R²"),
      caption = "Repeated 10-fold CV results for the linear regression model.")
```

The linear regression model produced consistent RMSE values across all 30 folds, with a low standard deviation indicating stable generalization. The mean R-squared is consistent with the adjusted R-squared reported during model fitting, confirming the model is not overfitting to the training data.

## Model 2 Prediction 

<<<<<<< HEAD
```{r echo = FALSE,warning=FALSE,message=FALSE,comment=""}
model2 <- train(salary ~ ., data=small.sample, method = "ranger", trControl= ctrl, num_trees= 50, tuneLength=1)
model2_results <- model2$results

kable(model2_results, digits = 2, caption = "Model 2 Training Results")



```


```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

small.baked <- baked.train[sample(nrow(baked.train), nrow(baked.train) * 0.1), ]

rf.cv <- train(
  salary ~ .,
  data      = small.baked,
  method    = "rf",
  trControl = cv.control,
  tuneGrid  = data.frame(mtry = 3),
  ntree     = 500
)

rf.cv.tbl <- data.frame(
  Model     = "Random Forest",
  Mean_RMSE = round(rf.cv$results$RMSE, 2),
  SD_RMSE   = round(rf.cv$results$RMSESD, 2),
  Mean_MAE  = round(rf.cv$results$MAE, 2),
  Mean_Rsq  = round(rf.cv$results$Rsquared, 4)
)

kable(rf.cv.tbl,
      col.names = c("Model", "Mean RMSE", "SD RMSE", "Mean MAE", "Mean R²"),
      caption = "Repeated 10-fold CV results for the random forest model.")
```


The random forest was evaluated on a 10% sample of the training data due to computational constraints. The mean RMSE is higher than the linear regression, and the larger standard deviation across folds reflects greater variability, likely a result of the reduced training sample rather than a fundamental weakness of the model.

```{r echo=FALSE}
cv.comp <- bind_rows(lm.cv.tbl, rf.cv.tbl)

kable(cv.comp,
      col.names = c("Model", "Mean RMSE", "SD RMSE", "Mean MAE", "Mean R²"),
      caption = "Cross-validation performance comparison for both models.")
```

```{r echo=FALSE,warning=FALSE,message=FALSE,comment=""}
pred1<- predict(model1, small.sample)
 
pred2<- predict(model2, small.sample)
 
prediction_results <- data.frame(

Actual_Salary = small.sample$salary,
 
Linear_Pred = pred1,
 
RF_Pred = pred2
 
)
 
```


## Model Comparison

<<<<<<< HEAD
The table below shows the five-number summaries of RMSE in both models. It clearly illustrates that Random forest has a higher RMSE in comparison with Linear regression. 
```{r echo=FALSE,warning=FALSE,message=FALSE,comment=""}
# 1. You MUST define 'results' first!
results <- resamples(list(Linear_Regression = model1, Random_Forest = model2))

# 2. Now you can extract the statistics
comparison_summary <- summary(results)$statistics$RMSE

# 3. Convert to data frame
comparison_table <- as.data.frame(comparison_summary)

# 4. Display with kable
knitr::kable(comparison_table[, 1:6], 
             digits = 2, 
             format.args = list(big.mark = ","),
             caption = "Comparison of RMSE Distribution")

```



The performance comparison highlights a significant advantage for the Linear Regression model. It achieved a substantially lower Mean RMSE compared to the Random Forest, indicating much higher predictive accuracy. Furthermore, the SD_RMSE and the five-number summary reveal that Linear Regression is more consistent; its maximum error is still considerably lower than the Random Forest's minimum error, proving it is the more stable model for this data set. 
```{r echo= FALSE,warning=FALSE,message=FALSE,comment=""}

# 1. Collect results
results <- resamples(list(Linear_Regression = model1, Random_Forest = model2))

# 2. Create the summary table
model_comp <- data.frame(
  Model = c("Linear Regression", "Random Forest"),
  Mean_RMSE = c(mean(model1$resample$RMSE), mean(model2$resample$RMSE)),
  SD_RMSE = c(sd(model1$resample$RMSE), sd(model2$resample$RMSE))
)

# 3. Display neatly
knitr::kable(model_comp, 
             digits = 2, 
             format.args = list(big.mark = ","),
             col.names = c("Model", "Mean RMSE (Accuracy)", "SD RMSE (Stability)"),
             caption = "Final Performance Comparison")
```





The following table and plot visualize the predictive performance of both models compared to actual salaries. On the plot, both the Linear Regression and Random Forest models track the actual salary trends closely; however, the Linear model exhibits higher variance in certain regions, occasionally underestimating the actual values more significantly than the Random Forest. 
```{r echo=FALSE,warning=FALSE,message=FALSE,comment=""}
library(ggplot2)
library(tidyr)
library(dplyr)

# Add an ID column for the x-axis (Observation Index)
prediction_results <- prediction_results %>%
  mutate(obs_id = row_number())

# Prepare data for plotting: pivot from wide to long format
plot_data <- prediction_results %>%
  pivot_longer(cols = c(Actual_Salary, Linear_Pred, RF_Pred), 
               names_to = "Type", 
               values_to = "Salary")
```

```{r echo=FALSE,out.width="65%", out.height="28%", fig.align='center', fig.cap="Salary Prediction Comparison",warning=FALSE,message=FALSE,comment=""}
# Plotting the first 50 observations for clarity
ggplot(plot_data %>% filter(obs_id <= 50), aes(x = obs_id, y = Salary, color = Type)) +
  geom_line(aes(linetype = Type), size = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("Actual_Salary" = "black", 
                                "Linear_Pred" = "blue", 
                                "RF_Pred" = "red")) +
  labs(title = "Salary Prediction Comparison: Actual vs. Models",
       subtitle = "First 50 observations from the 10% sample",
       x = "Observation Index",
       y = "Salary",
       color = "Legend",
       linetype = "Legend") +
  theme_minimal()
```

Below is the first 10 observations comparison of actual salary and model predictions
```{r echo=FALSE,warning=FALSE,message=FALSE,comment=""}
library(knitr)

# Simple table with comma formatting and 2 decimal places
kable(head(prediction_results, 10), 
      digits = 2, 
      format.args = list(big.mark = ","),
      caption = "Comparison of Actual vs. Predicted Salaries",
      col.names = c("Row ID", "Actual Salary", "Linear Pred", "RF Pred", "Obs ID"))


```


```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="RMSE per fold across repeated CV for both models."}
lm.fold.df <- data.frame(
  Fold  = seq_along(lm.cv$resample$RMSE),
  RMSE  = lm.cv$resample$RMSE,
  Model = "Linear Regression"
)

rf.fold.df <- data.frame(
  Fold  = seq_along(rf.cv$resample$RMSE),
  RMSE  = rf.cv$resample$RMSE,
  Model = "Random Forest"
)

fold.df <- bind_rows(lm.fold.df, rf.fold.df)

ggplot(fold.df, aes(x = Fold, y = RMSE, color = Model, group = Model)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(x = "Fold (10 folds × 3 repeats)",
       y = "RMSE",
       color = "Model") +
  theme_minimal()
```

The line plot shows the RMSE across all 30 folds for both models. The linear regression line remains flat and consistent, confirming stable generalization across different subsets of the data. The random forest shows more inconsistent variation, consistent with the higher standard deviation reported in the table. Overall, the linear regression outperforms the random forest under cross-validation, which mirrors the test-set findings reported earlier.

# References
[1] https://www.kaggle.com/datasets/nalisha/job-salary-prediction-dataset
