---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output: html_document:
    df_print: paged
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# Install these libraries if necessary (run in Console, not here)
library(knitr)
library(caret)
library(rpart)
library(pROC)
library(ggplot2)
library(rattle)
library(randomForest)
# reproducibility
rand.v = "4.3.2"
seed.val = 123456
```

# Overview
#Data Description and Processing
The dataset used in this analysis was come from Kaggle. We categorized the variables in the dataset as personal  and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.

The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?

Personal Variables: \
Experience years:Number of years of professional experience 
Education level:Highest level of education completed 
Skills count: Number of technical or professional skills 
Certifications Number of professional certifications


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer)
Industry: Industry sector where the job belongs 
Company size: Size of the company (small, medium, large)
Location:Job location or region 
Remote Work:Whether the job allows remote work 

Outcome Variable: 
Salary: Annual salary of the employee






```{r}
js.df<- read.csv("job_salary_prediction_dataset.csv")
sum(is.na(js.df))
```






