---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output: html_document:
    df_print: paged
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r echo=FALSE, include=FALSE, message=FALSE}
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)

# params for random generator
r.version <- "4.3.2"
seed.val <- 1234
```

# Overview


The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?


#Data Description and Processing
The dataset used in this analysis was come from Kaggle[1]. We categorized the variables in the dataset as personal and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.


Personal Variables: \
Experience years:Number of years of professional experience. Numeric \
Education level:Highest level of education completed. Categorical \
Skills count: Number of technical or professional skills. Numeric \ 
Certifications Number of professional certifications. Numeric \


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer). Categorical \
Industry: Industry sector where the job belongs. Categorical \
Company size: Size of the company (small, medium, large). Categorical \
Location:Job location or region. Categorical \
Remote Work:Whether the job allows remote work. Categorical \

Outcome Variable: 
Salary: Annual salary of the employee. Numeric \

The dataset was check for missing values. Of the original 250,000 observations, we found 0 observation that contained missing value. 



```{r}
js.df<- read.csv("job_salary_prediction_dataset.csv")
sum(is.na(js.df))
```


# Data Exploration 

In figure 1, the distribution of the outcome variable, salary, shows a mostly symmetrical shape that is mildly skewed to the right.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Salary distribution."}
hist(js.df$salary/1000000, freq = FALSE, # freq = FALSE for density scale on y-axis
     main = "", # turn off the title as using the fig.cap
     xlab = "Salary (in millions)",
     ylab = "Density",
     ylim = c(0, 11)
     )
lines(density(js.df$salary/1000000), col = "blue", lwd = 2)
```

#Modeling
##Model 1 

We fitted a linear regression model using all the predictors, with salary as the predicted variable.The estimates for the coefficients, standard errors and significance results are displayed below:
```{r echo=FALSE}
fit1 <- lm(salary~ .,data=js.df)
fit1.summary<- summary(fit1)
coef.vals <- coef(fit1.summary)
kable(coef.vals, caption="Summary of coefficients for model 1.")
vif1<- vif(fit1)
```






