---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output:
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r echo=FALSE, include=FALSE, message=FALSE}
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)
library(knitr) # for kable
library(caret) # creating test and training sets
library(rpart) # regression and classification trees
library(pROC) # roc plot
library(ggplot2)
library(rattle) #for fancyRpartPlot
library(randomForest)
# for fixing the random generator version and seed
#library(recipes)
#library(embed)
#library(tidymodels)
rand.v = "4.3.2"
seed.val = 123456
```

# Overview

There are many factors that determine the salary of a job. There are aspects that the job requires and qualities that the applicant can uniquely bring. Knowing how each of these predictors affect the salary of a job is of major interest to employers and those looking to be employed as they help them gauge the market.

The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?


#Data Description and Processing
The dataset used in this analysis was come from Kaggle[1]. We categorized the variables in the dataset as personal and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.


Personal Variables: \
Experience years:Number of years of professional experience. Numeric \
Education level:Highest level of education completed. Categorical \
Skills count: Number of technical or professional skills. Numeric \ 
Certifications Number of professional certifications. Numeric \


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer). Categorical \
Industry: Industry sector where the job belongs. Categorical \
Company size: Size of the company (small, medium, large). Categorical \
Location:Job location or region. Categorical \
Remote Work:Whether the job allows remote work. Categorical \

Outcome Variable: 
Salary: Annual salary of the employee. Numeric \

The dataset was check for missing values. Of the original 250,000 observations, we found 0 observation that contained missing value. 



```{r echo=FALSE}
js.df<- read.csv("job_salary_prediction_dataset.csv")
```


# Data Exploration 

In figure 1, the distribution of the outcome variable, salary, shows a mostly symmetrical shape that is mildly skewed to the right.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Salary distribution."}
hist(js.df$salary/1000000, freq = FALSE, # freq = FALSE for density scale on y-axis
     main = "", # turn off the title as using the fig.cap
     xlab = "Salary (in millions)",
     ylab = "Density",
     ylim = c(0, 11)
     )
lines(density(js.df$salary/1000000), col = "blue", lwd = 2)
```

Since the median is lower than the mean, the data is slightly right-skewed.

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$salary
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Table 1) Salary descriptive statistics")
```






The distribution of the histogram is a uniform distribution with a non-obvious outlier at the low end 
```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Experience Years."}
hist(js.df$experience_years, 
     freq = FALSE, 
     main = "Experience Year",
     yaxt = 'n', # Hide the default axis
     xlab = "Experience Year",
     ylab = "Percentage (%)",
     breaks = 20)

# Create a custom axis that multiplies density by 100
axis(2, at = seq(0, 0.1, by = 0.02), labels = seq(0, 10, by = 2))
lines(density(js.df$experience_years), col = "blue", lwd = 2)
```




Since the distribution of the graph is uniform distribution, the mean and median are nearly the same with a 0.01 difference. 
```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$experience_years
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Table 1) Salary descriptive statistics")
```



The distributino of the Job Title is uniform distributed, nearly all of the job title has the similar percentage. 
```{r echo=FALSE}
# 1. Create a frequency table
job_counts <- table(js.df$job_title)

# 2. Convert to percentages
job_percent <- prop.table(job_counts) * 100

# 3. Sort (optional)
job_percent <- sort(job_percent, decreasing = TRUE)

# 4. Create the bar plot
barplot(job_percent, 
        main = "Job Title Distribution", 
        ylab = "Frequency (%)",   # Labels the y-axis
        col = "skyblue", 
        las = 2, 
        cex.names = 0.7)
```



Because the distribution is uniformed, the frequency and percentage are nearly identical
```{r echo=FALSE}
# 1. Get raw counts
counts <- table(js.df$job_title)

# 2. Get percentages (rounded to 2 decimal places for neatness)
percentages <- round(prop.table(counts) * 100, 2)

# 3. Combine into a data frame
job_summary <- data.frame(
  Job_Level = names(counts),
  Frequency = as.numeric(counts),
  Percentage = as.numeric(percentages)
)

# 4. Sort by highest frequency
job_summary <- job_summary[order(-job_summary$Frequency), ]

# View the table
print(job_summary)
```

#Modeling

##Model 1 

```{r echo=false}
js.df <- js.df %>%
  mutate(continent = case_when(
    location %in% c("India", "Singapore") ~ "Asia",
    location %in% c("Australia") ~ "Oceania",
    location %in% c("USA", "Canada") ~ "North America",
    location %in% c("Germany", "Netherlands", "Sweden", "UK") ~ "Europe",
    location == "Remote" ~ "Remote",
    TRUE ~ "Other"
  )) %>%
  # Convert the new column to a factor
  mutate(continent = as.factor(continent))
```


We fitted a linear regression model using all the predictors, with salary as the predicted variable.The estimates for the coefficients, standard errors and significance results are displayed below:
```{r echo=FALSE}
salary_recipe <- recipe(salary ~ ., data = train.data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location, outcome = vars(salary)) %>%
  step_lencode_glm(industry, outcome = vars(salary)) %>%
  step_lencode_glm(education_level, outcome = vars(salary)) %>%
  step_normalize(all_numeric_predictors())

lm_spec <- linear_reg() %>% set_engine("lm")

salary_wf <- workflow() %>%
  add_recipe(salary_recipe) %>%
  add_model(lm_spec)

fit_lm <- fit(salary_wf, data = train.data)

lm.pred <- predict(fit_lm, test.data) %>%
  bind_cols(test.data)

lm.rmse <- sqrt(mean((lm.pred$.pred - lm.pred$salary)^2))
lm.rmse
```


```{r}
fit2 <- lm(salary~ experience_years+education_level+skills_count+certifications,data=js.df)
```

```{r}
fit3<- lm(salary~job_title+industry+company_size+location, data=js.df)
summary(fit3)
```



#model 2 
planning on using random forest, but not done

```{r echo=FALSE}

#RNGversion(rand.v)
#set.seed(seed.val)

#index.train <- createDataPartition(y = js.df$salary, p = 0.75, list = FALSE)
#train.data <- js.df[index.train,]
#test.data <- js.df[-index.train,]
```

```{r echo=FALSE}
#RNGversion(rand.v)
#set.seed(seed.val)
#small.sample <- train.data[sample(nrow(train.data), nrow(train.data) * 0.1), ]
#salary.df<-randomForest(salary ~., data=small.sample, mtry=3, importance=TRUE)'''
```
```


#References
[1] https://www.kaggle.com/datasets/nalisha/job-salary-prediction-dataset
