---
title: "INFO248-project2ms1-Wong-Wu"
author: "Zijian Wu & Adrian Wong"
date: "04/03/2026"
output:
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r echo=FALSE, include=FALSE, message=FALSE}
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)
library(knitr) # for kable
library(caret) # creating test and training sets
library(rpart) # regression and classification trees
library(pROC) # roc plot
library(ggplot2)
library(rattle) #for fancyRpartPlot
library(randomForest)
# for fixing the random generator version and seed
library(recipes)
library(embed)
library(tidymodels)
r.version <- "4.3.2"
seed.val = 123456
```

# Overview

There are many factors that determine the salary of a job. There are aspects that the job requires and qualities that the applicant can uniquely bring. Knowing how each of these predictors affect the salary of a job is of major interest to employers and those looking to be employed as they help them gauge the market.

The main research question of this analysis is how do personal and job factors contribute to a the predicted salary of a job?


#Data Description and Processing
The dataset used in this analysis was come from Kaggle[1]. We categorized the variables in the dataset as personal and job based. Knowing how these different factors affect possible job salaries is of major interest to both employers and applicants to gauge the market.


Personal Variables: \
Experience years:Number of years of professional experience. Numeric \
Education level:Highest level of education completed. Categorical \
Skills count: Number of technical or professional skills. Numeric \ 
Certifications Number of professional certifications. Numeric \


Job Variables:\
Job title: The job role or position (e.g.Data Analyst, AI engineer). Categorical \
Industry: Industry sector where the job belongs. Categorical \
Company size: Size of the company (small, medium, large). Categorical \
Location:Job location or region. Categorical \
Remote Work:Whether the job allows remote work. Categorical \

Outcome Variable: 
Salary: Annual salary of the employee. Numeric \

The dataset was check for missing values. Of the original 250,000 observations, we found 0 observation that contained missing value. 



```{r echo=FALSE}
js.df<- read.csv("job_salary_prediction_dataset.csv")
```


# Data Exploration 

In figure 1, the distribution of the outcome variable, salary, shows a mostly symmetrical shape that is mildly skewed to the right.

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Salary distribution."}
hist(js.df$salary/1000000, freq = FALSE, # freq = FALSE for density scale on y-axis
     main = "", # turn off the title as using the fig.cap
     xlab = "Salary (in millions)",
     ylab = "Density",
     ylim = c(0, 11)
     )
lines(density(js.df$salary/1000000), col = "blue", lwd = 2)
```

Since the median is lower than the mean, the data is slightly right-skewed.

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$salary
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Table 1) Salary descriptive statistics")
```






The distribution of the histogram is a uniform distribution with a non-obvious outlier at the low end 
```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Experience Years."}
hist(js.df$experience_years, 
     freq = FALSE, 
     main = "Experience Year",
     yaxt = 'n', # Hide the default axis
     xlab = "Experience Year",
     ylab = "Percentage (%)",
     breaks = 20)

# Create a custom axis that multiplies density by 100
axis(2, at = seq(0, 0.1, by = 0.02), labels = seq(0, 10, by = 2))
lines(density(js.df$experience_years), col = "blue", lwd = 2)
```




Since the distribution of the graph is uniform distribution, the mean and median are nearly the same with a 0.01 difference. 
```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "Mean","Median", "Max", "StdDev")
data<-js.df$experience_years
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,avg,md,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Table 1) Salary descriptive statistics")
```


According to the data, the mean salary for entry-level positions is approximately $30,000 less than that of experienced roles. This observation is based solely on years of experience and does not account for variables such as job title
```{r echo=FALSE}
# 1. Calculate and store the means
mean_entry <- mean(js.df$salary[js.df$experience_years == 0], na.rm = TRUE)
mean_exp   <- mean(js.df$salary[js.df$experience_years != 0], na.rm = TRUE)

# 2. Create a data frame for the table
comparison_table <- data.frame(
  Experience_Level = c("Entry Level (0 Years)", "Experienced (1+ Years)"),
  Average_Salary = c(mean_entry, mean_exp)
)

# 3. Generate the pretty table
kable(comparison_table, 
      col.names = c("Experience Group", "Mean Salary"),
      digits = 2,
      format.args = list(big.mark = ","),
      caption = "Salary Comparison: Entry Level vs. Experienced")
```



The line plot demonstrates a positive correlation between years of experience and salary. This upward trend indicates that as professional experience increases, compensation levels tend to rise accordingly
```{r echo=FALSE}
# Calculate means using aggregate
summary_base <- aggregate(salary ~ experience_years, data = js.df, mean)

# Plot
plot(summary_base$experience_years, summary_base$salary, 
     type = "b", # "b" means both lines and points
     col = "blue",
     xlab = "Experience", ylab = "Salary",
     main = "Salary Trend")
```



The distributino of the Job Title is uniform distributed, nearly all of the job title has the similar percentage. 
```{r echo=FALSE}
# 1. Create a frequency table
job_counts <- table(js.df$job_title)

# 2. Convert to percentages
job_percent <- prop.table(job_counts) * 100

# 3. Sort (optional)
job_percent <- sort(job_percent, decreasing = TRUE)

# 4. Create the bar plot
barplot(job_percent, 
        main = "Job Title Distribution", 
        ylab = "Frequency (%)",   # Labels the y-axis
        col = "skyblue", 
        las = 2, 
        cex.names = 0.7)
```


Because the distribution is uniformed, the frequency and percentage are nearly identical
```{r echo=FALSE}
# 1. Get raw counts
counts <- table(js.df$job_title)

# 2. Get percentages (rounded to 2 decimal places for neatness)
percentages <- round(prop.table(counts) * 100, 2)

# 3. Combine into a data frame
job_summary <- data.frame(
  Job_Level = names(counts),
  Frequency = as.numeric(counts),
  Percentage = as.numeric(percentages)
)

# 4. Sort by highest frequency
job_summary <- job_summary[order(-job_summary$Frequency), ]

kable(job_summary, col.names = c("Job Title", "Frequency", "Percentage (%)"),
      caption = "Table 2) Job title frequency and percentage.")
```

#Modeling

##Model 1 

We fitted a linear regression model using all the predictors, with salary as the predicted variable.The estimates for the coefficients, standard errors and significance results are displayed below:
```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

split      <- initial_split(js.df, prop = 0.75, strata = salary)
train.data <- training(split)
test.data  <- testing(split)

te.recipe <- recipe(salary ~ ., data = train.data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_lencode_glm(job_title, outcome = vars(salary)) %>%
  step_lencode_glm(location,  outcome = vars(salary)) %>%
  step_lencode_glm(industry,  outcome = vars(salary))

mod1     <- linear_reg(engine = "lm")
wf.te    <- workflow() %>% add_recipe(te.recipe) %>% add_model(mod1)
mod1.fit <- fit(wf.te, data = train.data)

prepped.rec <- extract_recipe(mod1.fit)
lm.engine   <- extract_fit_engine(mod1.fit)
```

```{r echo=FALSE}
coef.vals <- coef(summary(lm.engine))
kable(coef.vals, digits = 4,
      caption = "Table X) Model 1 coefficient estimates, standard errors, and significance.")
```

The target-encoded job title and location predictors ranked most important, each with a coefficient close to 1. This means that for every dollar increase in a category's average salary, the model predicts approximately one additional dollar in individual salary — confirming that the role and location are the most significant. Industry ranked near the bottom: the ten industry categories in this data encode to nearly identical average salaries, so the predictor contributes almost no salary signal to the model.

Among the directly interpreted predictors, company size showed the largest effects. Employees at startups earned approximately \$42,000 less on average than those at enterprise-sized companies, holding all other factors constant. Small and medium companies showed similar but smaller gaps. Education level showed meaningful differences, with PhD holders earning roughly \$21,000 more than those with a bachelor's degree. Experience years, skills count, and certifications all had positive but more modest effects.

The following tables show the average actual salary and average predicted salary for each level of the three target-encoded predictors. These allow the reader to interpret the encoded values in plain terms.

```{r echo=FALSE}
inspect.train.tbl <- train.data %>%
  transmute(salary, job_title, location, industry,
            experience_years, education_level, skills_count,
            certifications, company_size, remote_work) %>%
  bind_cols(
    bake(prepped.rec, new_data = train.data) %>%
      transmute(job_title_te = job_title,
                location_te  = location,
                industry_te  = industry),
    predict(mod1.fit, new_data = train.data)
  )
```

```{r echo=FALSE}
job.avg.tbl <- inspect.train.tbl %>%
  group_by(job_title) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),    big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),     big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(job_title_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(`Job Title` = job_title)

kable(job.avg.tbl,
      caption = "Table X) Average actual salary, predicted salary, and encoded value by job title.")
```

```{r echo=FALSE}
loc.avg.tbl <- inspect.train.tbl %>%
  group_by(location) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(location_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Location = location)

kable(loc.avg.tbl,
      caption = "Table X) Average actual salary, predicted salary, and encoded value by location.")
```

```{r echo=FALSE}
ind.avg.tbl <- inspect.train.tbl %>%
  group_by(industry) %>%
  summarize(
    `Avg Actual Salary`    = paste0("$", format(round(mean(salary)),   big.mark = ",")),
    `Avg Predicted Salary` = paste0("$", format(round(mean(.pred)),    big.mark = ",")),
    `Encoded Value`        = paste0("$", format(round(first(industry_te)), big.mark = ",")),
    .groups = "drop"
  ) %>%
  arrange(desc(`Avg Actual Salary`)) %>%
  rename(Industry = industry)

kable(ind.avg.tbl,
      caption = "Table X) Average actual salary, predicted salary, and encoded value by industry.")
```

### Evaluation of Model 1 Fit

We checked for multicollinearity among the predictors by calculating the variance inflation factor (VIF) for each.

```{r echo=FALSE}
vif.res <- as.data.frame(vif(lm.engine))
vif.res <- tibble::rownames_to_column(vif.res, var = "Predictor")
kable(vif.res, digits = 2,
      caption = "Table X) Generalized variance inflation factors (GVIF) for Model 1 predictors.")
```

All VIF values are small, indicating no meaningful multicollinearity among the predictors.

\newpage

```{r echo=FALSE, out.width="80%", out.height="35%", fig.align='center', fig.cap="Residuals vs. fitted values and Q-Q plot for Model 1."}
par(mfrow = c(1, 2))
plot(lm.engine, which = 1)
plot(lm.engine, which = 2)
par(mfrow = c(1, 1))
```

The residuals in figure X are evenly spread above and below zero across the fitted range with no obvious curvature, suggesting the linear assumption is reasonable. The Q-Q plot shows the residuals track the diagonal closely through the middle of the distribution, with only minor departures at the tails. We checked for heteroskedasticity as a precaution.

```{r echo=FALSE}
ct <- coeftest(lm.engine, vcov. = vcovHC(lm.engine, type = "HC3"))
kable(ct[, 1:4], digits = 4,
      caption = "Table X) Heteroskedasticity-robust (HC3) standard errors for Model 1.")
```

```{r echo=FALSE}
bpt <- bptest(lm.engine)
bps  <- round(as.numeric(bpt$statistic), 4)
degf <- as.numeric(bpt$parameter)
pv   <- round(as.numeric(bpt$p.value), 4)
bp.tbl <- tibble(bps, degf, pv)
kable(bp.tbl, col.names = c("BP Statistic", "df", "p-value"),
      caption = "Table X) Breusch-Pagan test results for Model 1.")
```

The significance levels of all predictors were unchanged after applying robust standard errors, and the Breusch-Pagan test returned a non-significant p-value. We found no evidence of heteroskedasticity and conclude that the model provides a valid basis for drawing inferences about the predictors' effects on salary.

#model 2 
planning on using random forest, but not done

```{r echo=FALSE}
RNGversion(r.version)
set.seed(seed.val)

small.sample <- train.data[sample(nrow(train.data), nrow(train.data) * 0.1), ]
salary.rf <- randomForest(salary ~ ., data = small.sample, mtry = 3, ntree = 500, importance = TRUE)
```

```{r echo=FALSE}
imp.vals <- importance(salary.rf, type = 1)  # type=1: %IncMSE
imp.df <- data.frame(
  Predictor  = rownames(imp.vals),
  Importance = round(imp.vals[, 1], 2)
)
imp.df <- imp.df[order(-imp.df$Importance), ]
imp.df$Rank <- seq_len(nrow(imp.df))
imp.df <- imp.df[, c("Rank", "Predictor", "Importance")]

kable(imp.df, row.names = FALSE,
      col.names = c("Rank", "Predictor", "% Increase in MSE"),
      caption = "Table X) Predictor importance for Model 2 (random forest), ranked by percent increase in mean squared error.")
```

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align='center', fig.cap="Predictor importance plot for Model 2 (random forest)."}
varImpPlot(salary.rf, type = 1, main = "")
```

Location ranked as the most important predictor, then experience years and company size. Job title was fourth. This is somewhat different from Model 1, where job title and location were tied at the top through their target-encoded coefficients. In the random forest, experience years and company size were more important because the model can capture non-linear salary increases across experience levels and threshold effects between company size categories.

Industry again ranked near the bottom, consistent with Model 1's finding that industry contributes very little salary signal in this data.

#References
[1] https://www.kaggle.com/datasets/nalisha/job-salary-prediction-dataset
